---
title: 'Regression Models: Course Project'
author: "rtaph"
date: "December 21, 2014"
output: pdf_document
---

******************************

### Executive Summary
The goal of this assignment is to find a relationship between a set of variables and miles per gallon (MPG) from the `mtcars` dataset. I particular, I am looking at the following two questions:

* Is an automatic or manual transmission better for MPG?
* What is the MPG difference between automatic and manual transmissions?


### Data Processing
I begin the analysis by loading libraries and setting a few global parameters:
  
```{r chunkOpts, message = FALSE, echo=FALSE}
  # load needed libraries, set global options, and working directory
  library(knitr)
  opts_chunk$set(echo = TRUE)       
  setwd("~/Documents/Courses/datasciencecoursera/RegModsP1/")
```

### Exploratory Data Analysis
We begin by loading the data:

```{r chunkLoadData, cache = TRUE, echo=FALSE}
  # load data
  data(mtcars)
```

### Model Fitting
My strategy for model selection is to

```{r chunkMod1, cache = TRUE, echo=FALSE}
  # Fit a regression model of mpg to automatic transmission
  mod1 = lm(mpg~am, data = mtcars)
  summary(mod1)
```
A review of the model summary (see appendices) reveals that it is statistically significant (*p* = `r summary(mod1)$p.value`).

```{r chunkMod2, cache = TRUE, echo=FALSE}
  # Fit a step-wise regression model
  mod2 <- lm(mpg~., data = mtcars)
  aicFormula <- step(mod2, direction = "backward", trace = FALSE)
  aicFormula$coefficients
```

#### Model 2

### Coefficient Interpretation

### Findings

Uncertainty in the models can be quantified with the confidence intervals.

# Appendices
### Appendix A: Exploratory Charts

```{r chunkApendixA1, cache = TRUE, echo=FALSE}
  pairs(mtcars, pch = 19)
```


### Appendix A: Model Diagnostic Plots
```{r chunkApendixB1, cache = TRUE, echo=FALSE}
  #
  par(mfrow = c(2,2))
  plot(mod1)
```