---
title: 'Regression Models: Course Project'
date: "December 21, 2014"
output: pdf_document
---

******************************
```{r chunkOpts, message = FALSE, echo=FALSE, autodep=TRUE}
  # load needed libraries, set global options, and working directory
  library(knitr)
  opts_chunk$set(echo = TRUE)       
  setwd("~/Documents/Courses/datasciencecoursera/RegModsP1/")
```

## Executive Summary
The goal of this assignment is to find a relationship between a set of variables and miles per gallon (MPG) from the `mtcars` dataset. I particular, I am seeking to answer the following two questions: (1) *Is an automatic or manual transmission better for MPG?* and (2) *What is the MPG difference between automatic and manual transmissions?* The analysis finds that on average, standard transmissions beat automatic transmission in vehicle performance. This difference is 7.24 miles to the gallon in a simple linear regression, and a smaller (but nonetheless significant) 2.94 miles to the gallon when other variables are accounted for.
  
## Exploratory Data Analysis

```{r chunkLoadData, cache = TRUE, echo=FALSE}
  # load data
  data(mtcars)
  mtcars$am = as.logical(mtcars$am)
```

I have pre-processed the data so that `am` is converted to a logical variable instead of a numeric one. My exploratory charts are presented in *Appendix A*. From initial inspection of these visualizations, there appears to be strong correlation between `mpg` and the `cyl`, `disp`, `wt` variables (the regression lines seem to fit the data well). 

## Model Fitting
### Model 1

I begin the analysis by fitting a simple linear regression model of mpg to one of the hypothesized covariates from the exploratory stage. In this case, I have chosen to model automatic transmission, `am`, as it is the major variable of interest to this study. The coefficients of this model are presented here:

```{r chunkMod1, cache = TRUE, echo=FALSE, autodep=TRUE}
  # Fit a regression model of mpg to automatic transmission
  mod1 = lm(mpg~am, data = mtcars)
  summary(mod1)$coefficients
```

The `amTRUE` variable above is a dummy variable for standard transmission. The interpretation of the coefficients is that standard transmission improves car performance by an average `r round(summary(mod1)$coefficients[2],2)` miles per gallon compared to automatic transmissions. A review of the model summary (see appendices) reveals that the t-test for $H_{0}: \beta_{am} = 0$ versus $H_{a}: \beta_{am} \neq 0$ is statistically signficant (*p* < .01). The `summary(mod1)` printout (not shown) further reveals that the `am` variable captures an estimated `r round(summary(mod1)$r.squared *100,0)`% of the variance in `mpg`.

However, there may be confounding variables that explain this covariance. For this we turn to a more complex model that explores all variables. 

### Model 2

A more complex model may shed light on the true nature of the relationship between `mpg` and `am`. Many other linear models could be fit using combinations of the 10 available predictors. My strategy for model selection will be to use `R`'s built-in backward stepwise selection using the Akaike Information Criterion (AIC). The summary printout of this model follows:

```{r chunkMod2, cache = TRUE, echo=FALSE, autodep=TRUE}
  # Fit a step-wise regression model
  mod2 <- lm(mpg~., data = mtcars)
  aicFormula <- step(mod2, direction = "backward", trace = FALSE)
  summary(aicFormula)
```

The step function has trimmed the 10 independent variables down to a list of 3 predictors that best explain the variation in `mpg`. These three variables are weight (`wt`), quarter-mile time (`qsec`), and the transmission type (`am`). Here, the coefficient interpretation is that holding all other variables fixed, standard transmissions only increase MPG performance by 2.93. Fortunately, although this covariation is not as strong as originally thought, it is still significant (p < .05) and the magnitude is the same direction. The residual sum of squares is also much lower in the multivariate model (`r round(summary(aicFormula)$r.squared *100,0)`% compared to `r round(summary(mod1)$r.squared *100,0)`%)

The diagnostic plots (*Appendix B*) reveal that there is no apparent heteroskedasticity or covariance in the standarized residuals. The Q-Q plot also fits the experimental quantiles along the identity line nicely, suggesting normality. These characteristics give us confidence that our parametic tests are appropriate and that our conclusions are robust.

### Findings

An analysis of variance reveals that the improvements the second model are highly significant (*p* < .01). The printout of the anova is presented:

```{r chunkCompMods, cache = TRUE, echo=FALSE, autodep=TRUE}
  print(anova(aicFormula, mod1))
```

Accordingly, it is apparent that there are substantial variables that confound the relationship between MPG and the transmission type. Once these aspects are accounted for, 
we conclude that:

 * a manual transmission is better for MPG; and
 * Having a standard transmission improves car performance by an average 2.94 mpg compared to an automatic transmission.


\newpage


# Appendices
### Appendix A: Exploratory Charts

```{r chunkApendixA1, cache = TRUE, echo=FALSE, autodep=TRUE}
  pairs(mtcars, panel = panel.smooth, lwd = 2, pch=".", cex = 3)
```

```{r chunkApendixA2, cache = TRUE, echo=FALSE, autodep=TRUE}
  library(car)
  scatterplotMatrix(~mpg+disp+drat+wt|cyl, data=mtcars,main="Exploratory Chart")
```


### Appendix B: Diagnostic Plots (Multivariate Model)
```{r chunkApendixB1, cache = TRUE, echo=FALSE, autodep=TRUE}
  # print the standard lm diagnostic plots
  par(mfrow = c(2,2))
  plot(aicFormula)
```